---
title: "Change In Per Hectare Crop Production Over Time"
author: 'Mohsin Ramay'
date: '`r Sys.Date()`'
output:
  html_document:
    number_sections: true
    fig_caption: true
    toc: true
    #fig_width: 7
    #fig_height: 4.5
    theme: cosmo
    highlight: tango
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  R.options = list,
  dpi = 300
)
```

# Objective
- In this notebook, I'll visualize the trends of crop yields in most populated countries over time. 
- I'll also visualize the population growth in the world and the populated countries. 
  - I worked on few countries in this notebook to optimally utilize the space. However, readers can use this code for the countries and crops of their choice.  
- Also, I'll model and visualize the change in crop yield over time in the selected countries.

# Setup

## Load libraries

```{r message=FALSE, warning=FALSE}
# Data science
library(tidyverse)
library(janitor)
library(DataExplorer)
library(skimr)

# Modeling
library(tidymodels)

# Visualization
library(ggrepel)

# Tables
library(gt)
```

## Upload data

- I used the data for crop production (yield) and population in countries provided on `github`.

```{r message=FALSE, warning=FALSE}
crop <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/key_crop_yields.csv")

land <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-01/land_use_vs_yield_change_in_cereal_production.csv")
```

# Exploratory data analysis

- Exploratory data analysis (EDA) helps in swiftly understanding the data structure and patterns in it. 

## Data structure

```{r}
skim(crop)
```


```{r}
skim(land)
```

- Both crop and land data have missing values. 

### Missing data

*Crop data*

```{r fig.width=8}
crop %>%
  plot_missing()
```

*Population data*

```{r fig.width=8}
land %>% 
  plot_missing()
```

- The data consists of 11 distinct crops produced worldwide. 

- The data contains information for 249 unique countries. 

## Population data

### Data distribution (2019)

```{r}
# Population in Millions
Pop <- land %>%
  filter(!is.na(Code)) %>%
  mutate(Year = as.numeric(Year)) %>% 
  mutate(pop = `Total population (Gapminder)`/1000000)

## Distribution (2019)
Pop %>% 
  filter(Entity != "World") %>%
  group_by(Entity) %>%
  filter(Year == max(Year)) %>%
  ungroup() %>% 
  ggplot(aes(x = pop)) +
  geom_histogram(fill = "steelblue") +
  labs(x = "Population", y = "Count") +
  theme_bw()
```

### Top 10 Popultated Countries

```{r}
Pop %>%
  filter(Entity != "World") %>%
  group_by(Entity) %>%
  filter(Year == max(Year)) %>%
  ungroup() %>% 
  slice_max(pop, n = 10) %>% 
  select(Entity, pop) %>%
  rename(Country = Entity, `Population (millions)` = pop) %>% 
  gt() %>% 
  tab_header(
    title = "Top 10 Populated Countried of the World"
  )
```


**Note:** - I'll use data of China, India, United States, Pakistan, Brazil and Russia for comparison. 


### Population Trends

#### Worldwide

```{r}
Pop %>% 
  filter(Entity == "World") %>% 
  filter(Year>=1900) %>% 
  ggplot(aes(x = Year, y = pop)) +
  geom_line() +
  geom_point() +
  labs(x= "Year", y = "Population (Millions)") +
  theme_bw()
```

#### Populated countries

```{r}
Pop %>% 
  filter(Entity %in% c("Pakistan", "United States", "India", "China", "Brazil", "Russia")) %>%
  filter(Year>=1900) %>%  
  ggplot(aes(x = Year, y = pop, color = Entity)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Population (millions)", color = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

## Crop data 

### Data cleaning

- Here, I'll gather the data in a tidy format for subsequent visualizations and data analysis. 
- Also, I'll clean the crop names to make them easily readable. 

```{r}
Crop <- crop %>% 
  gather(crop, yield, `Wheat (tonnes per hectare)`:`Bananas (tonnes per hectare)`, na.rm = FALSE, convert = TRUE) %>%
  mutate(crop = str_remove(crop, regex(" \\(tonnes per hectare\\)", ignore_case = TRUE)))

Crop
```

### Distribution of crop yields 

- Distribution of crop yield data obtained from 1961 to 2019.

```{r}
Crop %>% 
  ggplot(aes(x = yield)) +
  geom_histogram(fill = "steelblue") +
  theme_bw() +
  labs(x = "Yield") +
  facet_wrap(~crop, scales = "free")
```

### Jitter plots

```{r}
Crop %>% 
  ggplot(aes(x = reorder(crop, yield), y = yield, color = crop)) +
  geom_jitter(position = position_jitter(0.2), alpha = 0.1) +
  labs(x = "Crop", y = "Yield") +
  theme_bw() +
  theme(legend.position = "none")
```

### Yield trends 

```{r}
Crop %>% 
  group_by(Year, crop) %>% 
  summarise(yield = mean(yield, na.rm = TRUE)) %>% 
  ggplot(aes(x = Year, y = yield, color = crop)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Yield (tons per hectare)", color = NULL) +
  theme_bw()
  theme(legend.position = "top")
```

### Trends by Countries

```{r fig.height=6, fig.width=8}
`%nin%` = Negate(`%in%`)
  
Crop %>% 
  filter(Entity %in% c("Pakistan", "United States", "India", "China", "Brazil", "Russia")) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  group_by(Entity, Year, crop) %>% 
  summarise(yield = mean(yield, na.rm = TRUE)) %>% 
  ggplot(aes(x = Year, y = yield, color = Entity)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Yield (tons per hectare)", color = NULL) +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
  theme(legend.position = "top")
```

## Joint Trends 

- Now, I'll plot both the population growth and yield change over time in same graph.
- These plots will help in understanding that which crop yield (per hectare) has changed with rising population. 

**Data joining**

- I'll join the yield and population data for the six most populated countries. 

```{r}
Pop_Crop <- Pop %>% 
  filter(Entity %in% c("Pakistan", "United States", "India", "China", "Brazil", "Russia")) %>%
  filter(Year >= 1960) %>% 
  unite(ID, Entity, Year, sep = "_", remove = FALSE, na.rm = FALSE) %>% 
  left_join(
            Crop %>% 
              filter(Entity %in% c("Pakistan", "United States", "India", "China", "Brazil", "Russia")) %>%
              unite(ID, Entity, Year, sep = "_", remove = FALSE, na.rm = FALSE) %>% 
              select(-Entity, -Code, -Year),
            by = c("ID" = "ID")
  )
```

### By Country

#### China

```{r}
cols <- c("Yield" = "#08519c", "Population" = "red")

Pop_Crop %>% 
  filter(Entity == "China" & !is.na(crop)) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
   theme(legend.position = "top")
```


#### India

```{r}
Pop_Crop %>% 
  filter(Entity == "India" & !is.na(crop)) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
   theme(legend.position = "top")
```

#### United States

```{r}
Pop_Crop %>% 
  filter(Entity == "United States" & !is.na(crop)) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
   theme(legend.position = "top")
```

#### Pakistan

```{r}
Pop_Crop %>% 
  filter(Entity == "Pakistan" & !is.na(crop)) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
   theme(legend.position = "top")
```

#### Brazil

```{r}
Pop_Crop %>% 
  filter(Entity == "Brazil" & !is.na(crop)) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
   theme(legend.position = "top")
```

#### Russia

```{r}
Pop_Crop %>% 
  filter(Entity == "Russia" & !is.na(crop)) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans")) %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~crop, scales = "free") +
   theme(legend.position = "top")
```

### By Crop

#### Wheat

```{r fig.height=4, fig.width=8}
Pop_Crop %>% 
  filter(!is.na(crop)) %>% 
  filter(crop == "Wheat") %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~Entity, scales = "free") +
  theme(legend.position = "top")
```


#### Rice

```{r fig.height=4, fig.width=8}
Pop_Crop %>% 
  filter(!is.na(crop)) %>% 
  filter(crop == "Rice") %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~Entity, scales = "free") +
  theme(legend.position = "top")
```


#### Maize

```{r fig.height=4, fig.width=8}
Pop_Crop %>% 
  filter(!is.na(crop)) %>% 
  filter(crop == "Maize") %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~Entity, scales = "free") +
  theme(legend.position = "top")
```


#### Potatoes

```{r fig.height=4, fig.width=8}
Pop_Crop %>% 
  filter(!is.na(crop)) %>% 
  filter(crop == "Potatoes") %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~Entity, scales = "free") +
  theme(legend.position = "top")
```

#### Soybeans

```{r fig.height=4, fig.width=8}
Pop_Crop %>% 
  filter(!is.na(crop)) %>% 
  filter(crop == "Soybeans") %>% 
  ggplot(aes(x = Year)) +
  geom_line(aes(y = yield, colour = "Yield")) +
  geom_line(aes(y = 0.05*pop, color = "Population"), group = 1) +
  scale_y_continuous(sec.axis = sec_axis(~./0.05, name = "Population (millions)")) +
  scale_colour_manual("", values=cols) +
  labs(y = "Yield (tons per hectare)", color = "Legend") +
  theme_bw() +
  facet_wrap(~Entity, scales = "free") +
  theme(legend.position = "top")
```

**Note:** Readers can use this code to produce maps for other crops. 

# Modeling Crop Yield improvement

- Now, I'll use simple linear models to estimate the improvement in crop yield (tons per hectare ) per year.
- For this purpose, I'll use `tidymodels` library. 

## Crop yield change over time

- I'll obtain the slope coefficients for each crop in the populated countries. 
  - Slop coefficient = change in crop yield (tons per hectare ) per year. 

```{r}
slopes <- Pop_Crop %>% 
  select(-ID, -Code, -`Cereal yield index`, -`Change to land area used for cereal production since 1961`, -`Total population (Gapminder)`, -pop) %>% 
  filter(crop %nin% c("Cassava", "Cocoa beans") & !is.na(yield)) %>%
  nest(yields = c(Year, yield)) %>% 
  mutate(
    fit = map(yields, ~ lm(yield ~ Year, data = .x))
    ) %>% 
  mutate(tidied = map(fit, tidy)) %>%
  unnest(tidied) %>% 
  filter(term == "Year") %>%
  mutate(p.value = p.adjust(p.value)) %>% 
  select(-yields, -fit)

slopes
```

### Visualization of slope coefficients

- Now, I'll visualize the slope coefficients against p-values of null hypothesis. 
- This visualization will help us in understanding that which countries have demonstrated the improved yield over time in each crop. 

```{r fig.height=6, fig.width=8}
slopes %>%
  ggplot(aes(x = estimate, y = p.value, label = Entity)) +
  geom_vline(
    xintercept = 0, lty = 2,
    size = 1.5, alpha = 0.7, color = "gray50"
  ) +
  geom_point(aes(color = crop), alpha = 0.8, size = 2.5, show.legend = FALSE) +
  scale_y_log10() +
  facet_wrap(~crop) +
  geom_text_repel(size = 3) +
  theme_light() +
  theme(strip.text = element_text(size = 12)) +
  labs(x = "Change in yield (tons per hectare) per year", y = "P-value")
```

- The vertical line indicates `slope = 0` no change. 
- Countries on the right side of the dashed line have demonstrated the improved yield over time. 
- While, countries on the left side have demonstrated negative change in yield over time. 